\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-04-05 papers/methods/paper.tex
Updating likelihood math notation
Removing todo section
Updating paper from Bobs feedback

% Abstract of the paper
\begin{abstract}
We present a hierarchical Bayesian model for use in supernova cosmology which takes into account a full range of analysis systematics and selection effects. This advances the work of \citet{Rubin2015} by including additional sources of systematic uncertainty and improved treatment of Malmquist bias whilst increasing numerical efficiency. Our model is optimised for efficient fitting and has been validated on a combined simulated dataset of more than $250\,000$ supernovae. Simplified statistical simulations for ideal supernovae cosmology show no bias, and sophisticated SNANA simulations quantify the bias for Flat $w$CDM cosmology, with a matter prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$ included to allow rigorous determination of model bias in $w$. SNANA simulations can recover the input cosmology to within $0.27\sigma$, while taking into account many more effects than standard analyses.
\end{abstract}
% Bias in SNANA tests across intrinsic dispersion models is found to range from $-0.15\sigma$ to $+0.27\sigma$ in $w$ for datasets of comparable strength to the Dark Energy Survey three-year analysis.

@@

{\red Todo: get info on all used lowz surveys and introduce them properly. Help in terms of the right papers to cite would be appreciated.}

Almost two decades have passed since the discovery of the accelerating universe \citep{Riess1998, Perlmutter1999}. Since that time, the number of observed Type Ia supernovae (SN Ia) have increased by more than an order of magnitude thanks to modern surveys at both low redshift \citep{Bailey2008, Freedman2009, Hicken2009,  Contreras2010, Conley2011}, and higher redshift \citep{Astier2006, Wood-Vasey2007, Balland2009, Amanullah2010}. Cosmological analyses of these supernova samples \citep{Kowalski2008, Conley2011, Suzuki2012, Betoule2014, Rest2014} have been combined with complimentary probes of large scale structureand the CMB. For a recent review, see \citet{Huterer2018}. Despite these prodigious efforts, the nature of dark energy remains an unsolved mystery.

@@

Whilst supernova observations take the form of time-series photometric measurements of brightness in many photometric bands, most analyses do not work from these measurements. Instead, most techniques fit these observations (along with redshift) to a supernova model, with the most widely used being that of the empirical SALT2 model \citep{Guy2007, Guy2010}. This model is trained separately before fitting the supernovae light curves for cosmology \citep{Guy2010, Mosher2014}. The resulting output from the model is, for each supernova, a characterised amplitude $x_0$ (which can be converted into apparent magnitude $m_B = -2.5\log(x_0)$), a stretch term $x_1$ and colour term $c$, along with a covariance matrix describing the uncertainty on these summary statistics. As such, the product at the end is a (redshift-dependent) population of $m_B$, $x_1$ and $c$.

@@

\begin{align}
\mu_C &= 5 \log\left[ \frac{(1+z)r}{10} \right],\\
r& =\frac{c}{H_0} \int_0^z \frac{dz'}{\sqrt{ \Omega_m (1+z')^3 + \Omega_k (1 + z')^2 + \Omega_\Lambda (1+z')^{3(1+w)}}}, \label{eq:mu0}\\
\mu_{\rm obs} &= m_B + \alpha x_1 - \beta c - M_B. \label{eq:mu1}
\end{align}

The benefit this analysis methodology provides is speed -- for samples of hundreds of supernovae or less, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with several limitations. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of Gaussian uncertainty. Secondly, the method of creating a global covariance matrix relies on computing partial derivatives and thus any uncertainty estimated from this method loses all information about uncertainty correlation. Thirdly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae.

Selection efficiency, such as the well known Malmquist bias \citep{MalmquistK.G.1922} is accounted for by correcting data. Simulations following survey observational strategies and geometry are used to calculate the expected bias in distance modulus, which is then added onto the observational data. As these effects are not built into the likelihood, their influence on the error budget is not captured fully in the $\chi^2$ distribution, and any subtle correlations between cosmological or population parameters and the bias is lost.

@@

To try and escape the limitations of the traditional approaches, several recent methods have adopted Approximate Bayesian Computation, where supernova samples are forward modelled in parameter space and compared to observed distributions. \citet{Weyant2013} provides an introduction into ABC methods for supernova cosmology in the context of the SDSS-II results \citep{Sako2014} and Flat $\Lambda$CDM cosmology, whilst \citet{Jennings2016} demonstrates their \textit{superABC} method on simulated first season Dark Energy Survey samples, described in \citet{Kessler2015}. In both examples, the supernova simulation package SNANA \citep{Kessler2009a} is used to forward model the data at each point in parameter space.

By building the systematic uncertainties and selection effects into the simulation package, there is vastly more freedom in how to treat and model those effects. Data does not need to be corrected, analytic approximations do not need to be applied, we are free to incorporate algorithms that simply cannot be expressed in a tractable likelihood. This freedom comes with a cost -- computation. The classical $\chi^2$ method's most computationally expensive step in a fit is matrix inversion. For ABC methods, we must instead simulate an entire supernova population in its entirety -- drawing from underlying supernova populations, modelling light curves, applying selection effects, fitting light curves and applying data cuts. This is an intensive process. Luckily, efficient sampling algorithms that have walkers that rely on the Markov properties of groups instead of individual walkers, such as Ensemble sampling \citep{Foreman-Mackey2013} allow easy parallelisation of parameter fits and can help ease the computational burden.

@@

We tested two models against simulated supernova sets; $\beta(c) = \beta_0 + \beta_1 c$ and $\beta(z) = \beta_0 + \beta_1 z$. See Section \ref{sec:simdes} for details on simulation generation. We found for both models that non-zero values for $\beta_1$ are preferred (even with constant $\beta$ used in simulation) due to severe degeneracy with selection effects. This degeneracy resulted in a significant bias in recovered cosmology, and so in our final model we continue to adopt the constant $\alpha$ and $\beta$ found in traditional analyses. As such, our calculation of distance modulus $\mu$ mirrors that found in Equations \eqref{eq:mu0} and \eqref{eq:mu1}.

@@

\begin{align}
\mathcal{L} &= P({\rm data} | \theta, S). \label{eq:like}
\end{align}

@@

\begin{align}
\mathcal{L} &= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, \label{eq:main}
\end{align}

>>>>

% Abstract of the paper
\begin{abstract}
We present a hierarchical Bayesian model for use in supernova cosmology which takes into account a full range of analysis systematics and selection effects. This advances the work of \citet{Rubin2015} by including additional sources of systematic uncertainty and improved treatment of Malmquist bias whilst increasing numerical efficiency. Our model is optimised for efficient fitting and has been validated on a combined simulated dataset of more than $250\,000$ supernovae. Simplified statistical simulations for ideal supernovae cosmology show no bias, and sophisticated SNANA simulations of the combined low-redshift and DES 3-year spectroscopic survey sample quantify the bias for flat $w$CDM cosmology, with a matter prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$ included to allow rigorous determination of model bias in $w$. SNANA simulations can recover the input cosmology to within $0.27\sigma$, while taking into account many more effects than standard analyses.
\end{abstract}
% Bias in SNANA tests across intrinsic dispersion models is found to range from $-0.15\sigma$ to $+0.27\sigma$ in $w$ for datasets of comparable strength to the Dark Energy Survey three-year analysis.

@@

Almost two decades have passed since the discovery of the accelerating universe \citep{Riess1998, Perlmutter1999}. Since that time, the number of observed Type Ia supernovae (SN Ia) have increased by more than an order of magnitude thanks to modern surveys at both low redshift \citep{Bailey2008, Freedman2009, Hicken2009,  Contreras2010, Conley2011}, and higher redshift \citep{Astier2006, Wood-Vasey2007, Frieman2008, Balland2009, Amanullah2010, Sako2014}. Cosmological analyses of these supernova samples \citep{Kowalski2008, Kessler2009, Conley2011, Suzuki2012, Betoule2014, Rest2014} have been combined with complimentary probes of large scale structureand the CMB. For a recent review, see \citet{Huterer2018}. Despite these prodigious efforts, the nature of dark energy remains an unsolved mystery.

@@

Whilst supernova observations take the form of time-series photometric measurements of brightness in many photometric bands, most analyses do not work from these measurements. Instead, most techniques fit these observations (along with redshift) to a supernova model, with the most widely used being that of the empirical SALT2 model \citep{Guy2007, Guy2010}. This model is trained separately before fitting the supernovae light curves for cosmology \citep{Guy2010, Mosher2014}. The resulting output from the model is, for each supernova, a characterised amplitude $x_0$ (which can be converted into apparent magnitude $m_B = -2.5\log(x_0)$), a stretch term $x_1$ and colour term $c$, along with a covariance matrix describing the uncertainty on these summary statistics. As such, the product at the end is a (redshift-dependent) population of $m_B$, $x_1$ and $c$, where the width of this distribution is called the intrinsic dispersion of the supernova population.

@@

\begin{align}
\mu_C &= 5 \log\left[ \frac{(1+z)r}{10} \right],\\
r& =\frac{c}{H_0} \int_0^z \frac{dz'}{\sqrt{ \Omega_m (1+z')^3 + \Omega_k (1 + z')^2 + \Omega_\Lambda (1+z')^{3(1+w)}}}, \label{eq:mu0}
%\mu_{\rm obs} &= m_B + \alpha x_1 - \beta c - M_B. \label{eq:mu1}
\end{align}
where $r$ is the comoving distance for redshift $z$ given a specific cosmology and $H_0$ is the current value of Hubble's constant in $\kmsmpc$.

The benefit this analysis methodology provides is speed -- for samples of hundreds of supernovae or less, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with several limitations. Firstly, formulating a $\chi^2$ likelihood requires a loss of model flexibility by building into the model assumptions of Gaussian uncertainty. Secondly, the method of creating a global covariance matrix relies on computing partial derivatives and thus any uncertainty estimated from this method loses all information about uncertainty correlation. For example, correlation the underlying supernova colour population's mean and skewness is highly correlated, however this correlation is lost when determining population uncertainty using numerical derivatives of population permutations. Thirdly, the computational efficiency is dependent on inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae.

Selection efficiency, such as the well known Malmquist bias \citep{MalmquistK.G.1922} is accounted for by correcting data. Specifically, Malmquist bias is the result of losing the fainter tail of the supernova population at high redshift when supernovae increasingly fall below the detection threshold. An example of Malmquist bias is illustrated in Figure \ref{fig:malmquist}. Simulations following survey observational strategies and geometry are used to calculate the expected bias in distance modulus, which is then added onto the observational data. As these effects are not built into the likelihood, their influence on the error budget is not captured fully in the $\chi^2$ distribution, and any subtle correlations between cosmological or population parameters and the bias is lost.

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{malmquist.pdf}
	\end{center}
	\caption{An example of the effects of Malmquist bias. Here are shown 1000 simulated supernovae redshifts and distance modulus given fiducial cosmology. The simulated survey is magnitude limited, and all supernovae brighter than magnitude 24 are successfully observed (shown as blue dots), and all dimmer than 24th magnitude are not successfully observed (shown as red dots). By binning the supernovae along redshift, and taking the mean distance modulus of the supernovae in each bin, we can see that at higher redshift where Malmquist bias kicks in, the population mean drops and becomes bias. This source of bias must either be corrected by adjusting the data (such as subtracting the found bias) for by incorporating Malmquist bias explicitly in the cosmological model.}
	\label{fig:malmquist}
\end{figure}

@@

To try and escape the limitations of the traditional approaches, several recent methods have adopted Approximate Bayesian Computation, where supernova samples are forward modelled in parameter space and compared to observed distributions. \citet{Weyant2013} provides an introduction into ABC methods for supernova cosmology in the context of the SDSS-II results \citep{Sako2014} and flat $\Lambda$CDM cosmology, whilst \citet{Jennings2016} demonstrates their \textit{superABC} method on simulated first season Dark Energy Survey samples, described in \citet{Kessler2015}. In both examples, the supernova simulation package SNANA \citep{Kessler2009a} is used to forward model the data at each point in parameter space.

By building the systematic uncertainties and selection effects into the simulation package, there is more freedom in how to treat and model those effects. Data does not need to be corrected, analytic approximations do not need to be applied, we are free to incorporate algorithms that simply cannot be expressed in a tractable likelihood. This freedom comes with a cost -- computation. The classical $\chi^2$ method's most computationally expensive step in a fit is matrix inversion. For ABC methods, we must instead simulate an entire supernova population -- drawing from underlying supernova populations, modelling light curves, applying selection effects, fitting light curves and applying data cuts. This is an intensive process.

@@

We tested two models against simulated supernova sets; $\beta(c) = \beta_0 + \beta_1 c$ and $\beta(z) = \beta_0 + \beta_1 z$. See Section \ref{sec:simdes} for details on simulation generation. We found for both models that non-zero values for $\beta_1$ are preferred (even with constant $\beta$ used in simulation) due to severe degeneracy with selection effects. This degeneracy resulted in a significant bias in recovered cosmology, and so in our final model we continue to adopt the constant $\alpha$ and $\beta$ found in traditional analyses. As such, our calculation of distance modulus $\mu$ mirrors that found in Equation \eqref{eq:mu0}.

@@

\begin{align}
\mathcal{L}(\theta; {\rm data}) &= P({\rm data} | \theta, S). \label{eq:like}
\end{align}

@@

\begin{align}
\mathcal{L}(\theta; {\rm data}) &= \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, \label{eq:main}
\end{align}
